---
title: "Visualizing Covid-19 Data"
author: "Jiayun Dong"
date: "3/30/2020"
output: 
  # pdf_document
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(jsonlite)
library(dplyr)
library(ggplot2)
library(viridis)
library(directlabels)
library(tidyr)

setwd("/Users/jiayun/Documents/GitHub/Covid19-viz")
remove(list = ls())
```

# Visualizing Covid-19 Data

## TLDR

* Covid-19 has been spreading globally with over 700,000 people infected. Over 150,000 people are infected by Covid-19 in the US. Disease containment efforts are relatively more effective in some Asian countries and regions.
* The growth in the number of cases has become slower recently, which is more likely to be a sign of inadequate tests instead of the effectiveness of social distancing efforts.Positive rate of Covid-19 tests can be used to measure whether sufficient tests has been performed relative to the spread of the disease. It grows from 10% to 17% in the past 10 days.
* Dispite the variation in testing policies, we argue that rough comparisons in the severity of the spread could be made between states with similar numbers of tests per million population. We have found, for example, that the spread is more severe in NJ than DC, MI than IN and MA than VT.

## Introduction

The goal of this report is to visualize the spread of Covid-19 and to understand the current situation in terms of where we are in the dynamics of the spread. We begin by a brief overview of the spread of Covid-19 across the globe and in the US. We then move onto the closer scrutiny at the US data at state-level. In particular we visualize the trajectory of the spread across time for individual states as well as make some qualititive attempts to compare the state of the spread across states.

However, due to the different testing policies across the globe and across the states in the US, comparisons among the countries and states may be misleading. After all, no one can be confirmed to be infected with Covid-19 if no test is performed. As such, the use of the figures in this report as means of prediction is discouraged. Comparisons among countries and states should also be made cautiously.

In this report, we use the data from [Johns Hopkins dataset](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series) (also uploaded in the current repository) for the global overview and the data from [The COVID Tracking Project](https://covidtracking.com) for understanding the development in the US. We also use US states population data (uploaded in this repository) in order to understand the spread and testing of the virus from the per capita perspective. 

This report is written with the data as of March 30th, 2020.

## Overview of the Spread of Covid-19

```{r echo=FALSE}
# read JHU data
df_raw = as.data.frame(read.csv(file="time_series_covid19_confirmed_global.csv", head = TRUE, sep =","))

# data cleaning
    # separate Hubei from mainland China and name change of some observations
    dat_international = df_raw %>% 
      mutate(Country.Region = if_else(Province.State == "Hubei", "Hubei", as.character(Country.Region)),
             Country.Region = if_else(Province.State == "Hong Kong", "Hong Kong", as.character(Country.Region)),
             Country.Region = if_else(Province.State == "Macau", "Macao", as.character(Country.Region)),
             Country.Region = if_else(Country.Region == "Taiwan*", "Taiwan", as.character(Country.Region)),
             Country.Region = if_else(Country.Region == "China", "China ex Hubei", as.character(Country.Region)),
             Country.Region = if_else(Country.Region == "United Kingdom", "UK", as.character(Country.Region)),
             Country.Region = if_else(Country.Region == "Korea, South", "South Korea", as.character(Country.Region))) %>%
      filter(Country.Region != "Diamond Princess")
    
    # sum(is.na(dat_international))  # check for missing values
    
    # aggregate by country
    dat_international = dat_international %>%
      select(-c(Lat, Long, Province.State)) %>%
      group_by(Country.Region) %>% 
      summarise_all(funs(sum), na.rm = TRUE)
    
    # convert to tidy panel data
    dat_international = gather(dat_international, "date", "cases", 2:ncol(dat_international))
    
    # date format
    dat_international = dat_international %>% 
      group_by(Country.Region) %>%
      mutate(date_index = row_number(),
             date = as.Date("2020-01-22") + date_index - 1) %>%
      select(-date_index)
```

### Covid-19 is a Global Pandemic

Covid-19 began spreading in China around the beginning of 2020 and continued to spread globally with around `r paste0(round(sum(df_raw[,ncol(df_raw)])/1000), ",000")` people infected with the disease. The figure belows 

```{r echo=FALSE}
# global overview
global_overview = dat_international %>% group_by(date) %>% summarise(total = sum(cases))
    

ggplot(global_overview, aes(x = date, y = total)) + 
  geom_line() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(breaks = "15 days") +
  ggtitle("Number of People Infected with Covid-19 in the World") +
  theme_minimal() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
```

```{r echo=FALSE}
# number of countries reaching 100
    dat_international_100 = dat_international %>% 
      spread(date, cases) %>%
      mutate_if(is.numeric, function(x) as.numeric(x >= 100))
    dat_international_100 = data.frame("date" = as.Date(colnames(dat_international_100)[-1]), 
                              "number_of_countries" = colSums(dat_international_100[,-1]))

# number of countries reaching 1000
    dat_international_1000 = dat_international %>% 
      spread(date, cases) %>%
      mutate_if(is.numeric, function(x) as.numeric(x >= 1000))
    dat_international_1000 = data.frame("date" = as.Date(colnames(dat_international_1000)[-1]), 
                              "number_of_countries" = colSums(dat_international_1000[,-1]))
```

The disease has been spreading in `r length(unique(df_raw$Country.Region))` countries and regions. Currently, there are `r max(dat_international_100$number_of_countries)` countries and regions with over 100 confirmed cases of Covid-19, and `r max(dat_international_1000$number_of_countries)` countries and regions with over 1,000 confirmed cases. These numbers are growing rapidly. Most countries and regions have reached the 100 and 1000 thresholds around mid-March.

```{r echo=FALSE}
# number of countries reaching 100
    ggplot(dat_international_100, aes(x = date, y = number_of_countries)) +
      geom_line() +
      ggtitle("Number of Countries/Regions with Over 100 Infections") + 
      theme_minimal() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())

# number of countries reaching 1000
    ggplot(dat_international_1000, aes(x = date, y = number_of_countries)) +
      geom_line() +
      ggtitle("Number of Countries/Regions with Over 1000 Infections") + 
      theme_minimal() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
```

Among these countries and regions, the following are the hardest-hit places. ^[Notice that we treat Hubei Province and the rest of China separatedly in our report.]

```{r echo=FALSE}
# event study: reaching 100 cases is defined as an outbreak
df_event = dat_international %>% 
  group_by(Country.Region) %>%
  mutate(not_over_100 = if_else(cases <= 99, 1, 0),
         day_index = row_number(),
         event_date = sum(not_over_100) + 1) %>%
  filter(day_index >= event_date) %>% 
  mutate(day = day_index - event_date) %>%
  select(Country.Region, day, cases) %>%
  arrange(Country.Region, day)

# modify dates for Hubei: the data is available from 1/22. China's confirmed cases reached 100 at around 1/19
df_event[df_event$Country.Region == "Hubei", 2] = df_event[df_event$Country.Region == "Hubei", 2] + 3

# countries ranked by cases
df_top = df_event %>% group_by(Country.Region) %>% summarise(maxcases = max(cases)) %>% arrange(desc(maxcases))
names(df_top) = c("Country/Region", "Number of Confirmed Cases")
head(df_top, 10)
```

The following figure visualizes the trajectory of the spread of Covid-19 in some of the countries and regions. In particular, for each country or region, we define the date when the number of confirmed cases of Covid-19 reached 100 as the date of the "outbreak" and plot how the cumulative number of confirmed cases evolves after the outbreak.^[We set the outbreak of Hubei Provence as Jan 19. While data is not available from JHU before Jan 22, we have checked from [other data source](https://shimo.im/sheets/tyWrrrqppYVwQtCW/gVSL1) that the number of confirmed cases has reached 100 for Hubei on that day.]

It seems that China (both Hubei Province and the rest of China) and South Korea has controlled the spread with very limited daily growth in the number of confirmed cases roughly a month after their outbreaks (day 30). The disease is growing at a somewhat controlable rate after the outbreak in some of the other Asian countries and regions, including Japan, Singapore, Hong Kong and Taiwan. The spread in rest of the countries in the figure is concerning: the numbers of people infected are growing rapidly at a "close-to-exponential" rate.

```{r, echo=FALSE}
# visualization parameter
xmax = ceiling(max(df_event$day)/5) * 5
ymax = max(df_event$cases) *1.05
country_list = c("China ex Hubei", "Hubei", "South Korea", "Iran", "Italy", "US", "Japan", "Singapore", "UK", "Germany", "Hong Kong", "Spain", "France", "Canada", "Taiwan", "Australia")
recent_date = as.Date(as.POSIXlt(Sys.time(), tz = "UTC")) - 1

# adjust no update issue on 3/13
df_event = df_event %>%
  mutate(cases = if_else(Country.Region %in% country_list & cases - lag(cases) <= 2 & !is.na(lead(cases)) & !is.na(lag(cases)),
                         (as.double(lag(cases)) * as.double(lead(cases)))^0.5, as.double(cases)))

ggplot(data = df_event %>% filter(Country.Region %in% country_list), aes(x = day, y = cases)) +
  geom_line(aes(color = Country.Region), size = 0.75) + 
  # geom_vline(xintercept = 0, linetype = "dashed", color = "gray", size=0.55) +
  scale_y_log10(breaks = c(100, 500, 1000, 5000, 10000, 50000, 100000, 150000), 
                minor_breaks = c(seq(from = 0, to = 1000, length.out = 5), seq(from = 0, to = 10000, length.out = 5),
                                 seq(from = 0, to = 100000, length.out = 5)), 
                labels = scales::comma,
                limits = c(100, ymax)) +
  scale_x_continuous(breaks = seq(from = 0, to = xmax, by = 5), limits = c(0, xmax)) +
  xlab(paste0("Days since confirmed cases reached 100 \n (JHU Data, available from 2020-01-22, updated at ", recent_date, " 23:59 UTC)")) + 
  ylab("Number of confirmed cases") +
  geom_dl(aes(label = Country.Region), method = list(dl.trans(x = x - 0.1), "last.points", cex = 0.8)) +
  theme_minimal() + theme(legend.position = "bottom")
```

We emphasize again that the number of confirmed cases in each country is affected by its testing policy and capacity, which are not uniform across the world. Moreover, countries with similar total number of confirmed cases may differ in the severity of the spread of Covid-19, due to the difference in total population. As such, while it is tempting to argue a country is more efficient in containing the virus than another country, this argument is unfortunately inaccurate.

### Spread of Covid-19 in the US

```{r echo=FALSE}

remove(list = ls())
# read COVID-TRACKING data
dat = as.data.frame(fromJSON("https://covidtracking.com/api/states/daily")) %>%  
  select(state, date, positive, negative, pending, death, total) %>%
  mutate(date = as.Date(as.character(date), format = "%Y%m%d"),
         death = if_else(is.na(death), 0, as.double(death)),
         pending = if_else(is.na(pending), 0, as.double(pending)),
         negative = if_else(is.na(negative), 0, as.double(negative)),
         totalTestResults = total - pending) %>%  # excluding tests with pending results
  arrange(state, date) %>%
  select(-total)

# read US population data
dat_pop = read.csv(file="StatePopulations.csv", head = TRUE, sep =",") %>% 
  mutate(state_abbrev = as.character(state_abbrev)) %>%
  select(state = state_abbrev, population = X2018.Population)

# join datafreames
dat = dat %>%
  left_join(dat_pop, by = c("state" = "state")) %>%
  mutate(positive_per_million_pop = round(positive / population * 1000000),
         death_per_million_pop = round(death / population * 1000000),
         test_per_million_pop = round(totalTestResults / population * 1000000))
```

Now we move onto US data. Currently, there are about `r paste0(round(dat %>% filter(date == max(date)) %>% summarise(total_cases = sum(positive, na.rm = TRUE)) %>% pull()/1000),",000")` confirmed cases of Covid-19 in the US. 

The figure below show the dynamics of the spread in the US. The situation worsens extremely rapidly in late-March, with the number of cases doubling every 3 days. It also shows the number of incremental cases reported after the outbreak of Covid-19 in the US. Recently, the number of cases grows by about 20,000 everyday.

Massive social distancing practices as well as stay-at-home order have been implemented recently across the nation. While it might still be too early to carefully analyze their effectiveness, we estimated the average growth rate in the timeframe between 3 and 7 days from today and use the estimated rate to "project" the number of cases for the recent 3 days. A lower actual number of cases than the projected number **might** indicate a slower growth rate in the most recent days. One can see from the figure that the "projected" trajectory (grey dashed curve) indeed lies above the actual trajectory. 

```{r echo=FALSE}
# the US trajectory of cases
  overview = dat %>%
    group_by(date) %>%
    summarise_if(is.numeric, sum, na.rm=TRUE) %>% # aggregate numeriac variables at national level
    mutate(positiveDailyIncrease = positive - lag(positive),
           testResultsDailyIncrease = totalTestResults - lag(totalTestResults),
           positiveDailyIncrease = replace_na(positiveDailyIncrease, 0), # the first day has no lag: set as zero
           testResultsDailyIncrease = replace_na(testResultsDailyIncrease, 0),
           positiveRate = positive / totalTestResults,
           day = row_number())

  # projection assuming same growth rate
    xmax = max(overview$day)
    ymax = max(overview$positive)
    y = overview$positive[(xmax - 7) : (xmax - 3)]
    x = overview$day[(xmax - 7) : (xmax - 3)]
    f = lm(log(y) ~ x)$coefficients
    l = function(x) exp(f[1] + f[2] * x)
    projection <- data.frame(x = (xmax - 7) : xmax, y = NA)
    projection$y <- l(projection$x)
  
  # time series of positive cases 
  ggplot(data = overview, aes(x = day)) +
    geom_line(aes(y = positive, color = "Positive Cases")) +
    geom_bar(stat = "identity", aes(y = positiveDailyIncrease, fill = "Daily Increase in Positive Cases")) +
    geom_line(data = projection, aes(x = x, y = y, color = "Projected Trajectory"), linetype = "dashed", show.legend=FALSE) +
    scale_x_continuous(breaks = seq(1,xmax, 5), labels = overview$date[seq(1,xmax, 5)]) +
    scale_y_continuous(labels = scales::comma) +
    scale_color_manual(values = c("darkred", "grey")) +
    scale_fill_manual(values = "darkblue") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.title = element_blank(), 
          axis.title.x = element_blank(), axis.title.y = element_blank()) +
    ggtitle("Overview of Covid-19 in the US: Positive Cases")
```


However, one should not be too optimistic about the current situation after observing this pattern. It is equally likely that the pattern is attributed to the inadequate growth in the testing capacity. As we argued before, the number of cases are constrained by the testing capacity. 

The figure below shows the cumulative number of Covid-19 tests performed in the US from the outbreak, together with its daily increments.^[Tests with pending results are excluded.] The ability to test Covid-19 infection is foundamental in containing the virus, as carriers of the virus may be asymptomatic. One can roughly see that the number of tests grows at the exponential rate from mid-March, followed by the linear rate in the few week. Therefore, attributing the slower growth rate in the spread of disease to effective social distancing measure might be unwarrented. 

```{r, echo=FALSE}
  # # projection assuming same growth rate
  #   xmax = max(overview$day)
  #   ymax = max(overview$totalTestResults)
  #   y = overview$totalTestResults[(xmax - 7) : (xmax - 3)]
  #   x = overview$day[(xmax - 7) : (xmax - 3)]
  #   f = lm(log(y) ~ x)$coefficients
  #   l = function(x) exp(f[1] + f[2] * x)
  #   projection <- data.frame(x = (xmax - 7) : xmax, y = NA)
  #   projection$y <- l(projection$x)

  # time series of testing
  ggplot(data = overview, aes(x = day)) +
    geom_line(aes(y = totalTestResults, color = "Total Test Results")) +
    geom_bar(stat = "identity", aes(y = testResultsDailyIncrease, fill = "Daily Increase in Test Results")) +
    # geom_line(data = projection, aes(x = x, y = y, color = "Projected Trajectory"), linetype = "dashed") +
    scale_x_continuous(breaks = seq(1,xmax, 5), labels = overview$date[seq(1,xmax, 5)]) +
    scale_y_continuous(labels = scales::comma) +
    # scale_color_manual(values = c("grey", "darkred")) +
    scale_color_manual(values = c("darkred")) +
    scale_fill_manual(values = "darkblue") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.title = element_blank(), 
          axis.title.x = element_blank(), axis.title.y = element_blank()) +
    ggtitle("Overview of Covid-19 in the US: Tests")
```

One metric, while it is not a perfect one, to measure if there are enough tests performed relative to the severity of the spread of disease is the positive rate, which is the percentage of people with positive test result among those tested for the virus. Persumably, only those who have symptoms similar to Covid-19 and the close contacts of infected people are tested for the virus. Fixing the relative sizes of these two groups, positive rate is higher when tests are reserved for people with more severe symptioms or when contact tracing is performed less agressively. Therefore, higher positive rate suggests insufficient testing capacity relative to the spread of the disease.

The figure below shows how the positive rate varies in the US after the outbreak. It starts from about 15% and decreases to the lowest level of 10% in mid-March. After that, the positive rate grows up to the current level of about 17%. This suggests that the testing capacity of the US is still inadequate relative to the spread of Covid-19 in the past 10 days. It also suggests that the slower growth rate in the number of cases in the recent days, which we noted earlier, may not be attributed to the effectiveness of social distancing efforts. At least there is not enough evidence to make this statement.

```{r echo=FALSE}
# time series of positive rate
  ggplot(data = overview, aes(x = day)) +
    geom_line(aes(y = positiveRate, color = "Positive Rate"), show.legend = FALSE) +
    scale_x_continuous(breaks = seq(1,xmax, 5), labels = overview$date[seq(1,xmax, 5)]) +
    scale_y_continuous(labels = scales::percent, limits = c(min(overview$positiveRate)*0.9, max(overview$positiveRate)*1.1)) +
    scale_color_manual(values = c("darkred")) +
    theme_minimal() +
    theme(legend.position = "bottom", legend.title = element_blank(), 
          axis.title.x = element_blank(), axis.title.y = element_blank()) +
    ggtitle("Overview of Covid-19 in the US: Positive Rate")
```

## US State-Level Analysis

```{r echo=FALSE}
# trajectories of states
df_trajectory = dat %>%
  group_by(state) %>%
  mutate(less_than_100 = if_else(positive <= 99, 1, 0),
         day = row_number(),
         outbreak = sum(less_than_100) + 1,
         day = day - outbreak) %>%
  filter(day >= 0) %>%
  select(state, day, positive, positive_per_million_pop, test_per_million_pop, totalTestResults)

# current situation
df_current = dat %>%
  group_by(state) %>% filter(date == max(date)) %>% arrange(desc(positive))

```

We visualize the trajectory of the spread of Covid-19 for the top 10 states in terms of total number of cases in the figure below. Observe from the figure that WA, NY and CA are the first three states where the outbreak started: The outbreak occurred more than 20 days ago in these state. NY, which is the state with the most cases of Covid-19, has `r max(df_current$positive, na.rm = TRUE)` people contracting the virus. Covid-19 has been spreading rapidly in NJ and MI: No other state has seen more cases of Covid-19 than NJ and MI ten days after the outbreak.

```{r echo=FALSE}

state_list = df_current %>% arrange(desc(positive)) %>% pull(state)   # interested in the top 10 states in terms of number of cases
state_list = state_list[1:10]
ymax = ceiling(max(df_trajectory$positive) / 1000) * 1000
xmax = ceiling(max(df_trajectory$day) / 5) *5

ggplot(data = df_trajectory %>% filter(state %in% state_list), aes(x = day, y = positive)) +
  geom_line(aes(color = state), size = 0.5) + 
  scale_y_log10(minor_breaks = c(seq(from = 100, to = 1000, length.out = 6), 
                                 seq(from = 1000, to = 10000, length.out = 6),
                                 seq(from = 10000, to = 100000, length.out = 6)),
                limits = c(100, 100000),
                labels = scales::comma) +
  scale_x_continuous(minor_breaks = seq(from = 0, to = xmax, by = 1)) +
  xlab(paste0("Days since cases reached 100")) + 
  ylab("Number of cases (in log scale)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "bottom") +
  theme_minimal() + 
  ggtitle("State Trajectories: Number of Cases") +
  geom_dl(aes(label = state), method = list(dl.trans(x = x - 0.1), "last.points", cex = 0.8)) +
  theme(legend.position = "bottom")
```

We study the trajectories of the following states in more details.

```{r echo=FALSE}

plot_state = function(state_name){
  df_trajectoy_state = df_trajectory %>% 
    filter(state == state_name) %>%
    mutate(positiveDailyIncrease = positive - lag(positive),
           positiveDailyIncrease = if_else(is.na(positiveDailyIncrease), 0, as.double(positiveDailyIncrease)),
           positiveRate = positive / totalTestResults)
  
  # projection assuming same growth rate
  xmax = max(df_trajectoy_state$day)
  ymax = max(df_trajectoy_state$positive)
  y = df_trajectoy_state$positive[(xmax - 7) : (xmax - 3)]
  x = df_trajectoy_state$day[(xmax - 7) : (xmax - 3)]
  f = lm(log(y) ~ x)$coefficients
  l = function(x) exp(f[1] + f[2] * x)
  projection <- data.frame(x = (xmax - 7) : xmax, y = NA)
  projection$y <- l(projection$x)
  
  # time series of positive cases 
  ggplot(data = df_trajectoy_state, aes(x = day)) +
    geom_line(aes(y = positive, color = "Positive Cases")) +
    geom_bar(stat = "identity", aes(y = positiveDailyIncrease, fill = "Daily Increase in Positive Cases")) +
    geom_line(data = projection, aes(x = x, y = y, color = "Projected Trajectory"), linetype = "dashed") +
    geom_line(aes(y = positiveRate * ymax, color = "Positive Rate")) +
    scale_y_continuous(labels = scales::comma, 
                       sec.axis = sec_axis(~. / ymax, name = "Positive Rate", 
                                           breaks = seq(0,1,length.out = 6),
                                           labels = c("0%", "20%", "40%", "60%", "80%", "100%"))) +
    scale_x_continuous(breaks = seq(0, xmax, 1)) +
    scale_color_manual(values = c("darkred", "red","grey")) +
    scale_fill_manual(values = c("darkblue")) +
    xlab("Days since cases reached 100") + ylab("Number of Positive Cases") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.title = element_blank(), 
          panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    ggtitle(paste0("Spread of Covid-19 in ", state_name))
}

```

#### New York

New York is the state that is hit the hardest by Coronavirus. The figure below displays the dynamics of the spread of Covid-19 in NY after its outbreak. The dark red curve represents the number of positive cases while the grey dashed curve is the projected number based on the average growth rate in a 5-day window between 3 and 7 days ago. The dark blue bars captures the daily increments in the number of positive cases. The light red curve represents the trajectory of NY's positive rate. Recall that we have argued that the positive rate of Covid-19 tests measures the testing capacity relative to the severity of the spread. 

The situation in NY is qualitively similar to the overall situation in the US. This is natural since roughly `r paste0(round(df_current %>% filter(state == "NY") %>% pull(positive) / df_current %>% ungroup() %>% summarize(total = sum(positive, na.rm = TRUE)) %>% pull() * 100), "%")` of the positive cases are contributed by NY. In US, the growth rate seems to become slower in recent days. Again, this cannot be attributed to the effectiveness of social distancing policies: It could be alternatively casued by the limited testing capaicity. Indeed, the positive rate in NY doubles from 10% to 20% in the past 10 days, indicating that tests are insufficient relative to the spread of the disease.

```{r echo=FALSE}
plot_state("NY")
```

#### California
The situation in CA might be even worse. Testing capacity is not growing fast enough: Positive rate of Covid-19 test rises from about 10% to over 20% in the past 10 days. Despite the constraining testing capacity, the growth rate in the number of positive cases accelerates, with the actual trajectory above the projected trajectory based on the average growth rate in the past few days.

```{r echo=FALSE}
plot_state("CA")
```

#### Massachusett

There could be good news in MA. The curve of positive rate suggests that testing capacity is not becoming more constraining. Yet, the growth rate in the number of positive cases decreases. This might suggest that social distancing measures are effective in MA.

```{r echo=FALSE}
plot_state("MA")
```

### Testing of Covid-19 at State Level

Testing of Covid-19 is essential since it is the first step to isolating the infected people and containing the virus. We have already seen a rapid growthing in the number of tests implemented in the US, and we break it down at the state level. The following two figures shows the number of tests and the number of tests per million population performed in the top 10 states according to the metrics, respectively. While we have argued that testing capacity is not sufficient relative to the severity of spread, we also want to point out that the states are making great effort to catch up.

```{r echo=FALSE}
state_list = df_current %>% arrange(desc(totalTestResults)) %>% pull(state)   # interested in the top 10 states in terms of total tests
state_list = state_list[1:10]
ymax = ceiling(max(df_trajectory$totalTestResults) / 1000) * 1000
xmax = ceiling(max(df_trajectory$day) / 5) * 5

ggplot(data = df_trajectory %>% filter(state %in% state_list), aes(x = day, y = totalTestResults)) +
  geom_line(aes(color = state), size = 0.5) + 
  scale_x_continuous(minor_breaks = seq(from = 0, to = xmax, by = 1)) +
  scale_y_continuous(labels = scales::comma) +
  xlab(paste0("Days since cases reached 100")) + 
  ylab("Number of tests ") + 
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  ggtitle("State Trajectories: Number of Tests") +
  geom_dl(aes(label = state), method = list(dl.trans(x = x - 0.1), "last.points", cex = 0.8))
```


```{r echo=FALSE}
state_list = df_current %>% arrange(desc(test_per_million_pop)) %>% pull(state)   # interested in the top 10 states in terms of test_per_million_pop
state_list = state_list[1:10]
ymax = ceiling(max(df_trajectory$test_per_million_pop) / 1000) * 1000
xmax = ceiling(max(df_trajectory$day) / 5) * 5

ggplot(data = df_trajectory %>% filter(state %in% state_list), aes(x = day, y = test_per_million_pop)) +
  geom_line(aes(color = state), size = 0.5) + 
  scale_x_continuous(minor_breaks = seq(from = 0, to = xmax, by = 1)) +
  scale_y_continuous(labels = scales::comma) +
  xlab(paste0("Days since cases reached 100")) + 
  ylab("Number of tests per million population ") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "bottom") +
  theme_minimal() + 
  ggtitle("State Trajectories: Number of Tests per Million Population") +
  geom_dl(aes(label = state), method = list(dl.trans(x = x - 0.1), "last.points", cex = 0.8)) +
  theme(legend.position = "bottom")
```

### Cross-State Comparison

We have used another metric to measure the adequacy of testing capacity in the analysis above, which is the number of tests per million population. It measures whether the number of tests performed is sufficient relative to the population. The advantage of this metric, compared with the positive rate, is that it is not affected by the testing policies that varies across states. However, the disadvantage is that it is not linked to the scale of the spread of the disease: We need more tests when the spread is more severe. 

We propose the following way to make cross-state comparison, which is demonstrated by the figure below. We conjecture that each state prioritizes testing those who are more likely to be infected by Covid-19, including symptomatic patients as well as close contacts of infected people. When two states have performed the same number of tests per million population, their tests should cover the groups of people who are equaly likely to be infected. Therefore, the comparison of the positive rate of Covid-19 tests between such two states should coincide with the relative severities of the spreads among the people with high risk, which is closely related to the relative severities of the spreads in the states.

Based on this logic and the figure below, we can argue, for example, that the spread is more severe in NY than WA, that it is more severe in NJ than DC, and that it is more severe in MI than IN.

```{r echo=FALSE}
df_test = df_current %>%
  mutate(pos_rate = positive / totalTestResults) %>%
  select(state, positive, totalTestResults, pos_rate, test_per_million_pop) %>%
  filter(positive >= 100) %>%
  arrange(desc(positive))

ggplot(data = df_test, aes(x = test_per_million_pop, y = pos_rate, label = state)) +
  geom_text(cex = 2.5, color = "black") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::comma) +
  xlab("Tests per million population") + ylab("Positive rate") + ggtitle("Cross-State Comparison") +
  theme_minimal() + theme(legend.position = "right")
  
```

### One Figure with Rich Information

We finally provide a figure that converys the latests Covid-19 data of each state in the US. The dashed line represents the national average positive rate: about 17 people are infected by the Coronavirus our of every hundred people tested. Each "bubble" in the figure represents a state and is placed according to the number of tests and number of cases of Covid-19 (in log-scale). The color of a "bubble" captures the positive rate of the tests performed by the state and the size of a "bubble" captures the number of tests performed per million population. We only visualize the data for a selected set of states given the limited size of figure, but more states could be added to the figure. 

```{r echo=FALSE}
df_test = df_current %>%
  mutate(pos_rate = positive / totalTestResults) %>%
  select(state, positive, totalTestResults, pos_rate, test_per_million_pop) %>%
  filter(positive >= 100) %>%
  arrange(desc(positive))

national_avg = df_current %>% 
  ungroup() %>%
  summarise(avg_positive_rate = sum(positive, na.rm = TRUE) / sum(totalTestResults, na.rm = TRUE),
            avg_test_per_million_pop = round(sum(totalTestResults, na.rm = TRUE) / sum(population, na.rm = TRUE) * 1000000))

slope = as.numeric(national_avg[1])
l = function(x) log10(slope * x)

title = paste0("COVID19 Testing in Selected States \nNational Average: Positive Rate: ", 
               round(national_avg[1] * 1000) / 10, 
               "%; Tests per Million Population: ", 
               round(national_avg[2]))

state_list = c("NY", "WA", "CA", "NJ", "NC", "DC", "DE", "MI") # these are the states with labels
state_bubble = union(unique(df_test$state)[sample.int(50,15, replace = FALSE)], state_list)

ggplot(df_test %>% filter(state %in% state_bubble), 
       aes(x = totalTestResults, y = positive, size = test_per_million_pop, color = pos_rate, label = state)) +
  geom_point(alpha = 0.4) +
  geom_text(cex = 2.5, color = "black") +
  geom_label(data = df_test %>% filter(state %in% state_list), 
             aes(label = paste0(round(pos_rate * 100), "%; ", test_per_million_pop)),
             cex = 2, color = "black", nudge_y = 0.15, alpha = 0.05) +
  scale_color_gradient(low = "yellow2", high = "red2", name = "Positive Rate", limits = c(0, 0.5), oob = scales::squish) +
  scale_size(range = c(3, 18), breaks = c(250, 1000, 4000, 8000), name = "Test Per Million Pop") +
  stat_function(fun = l, linetype = "dashed", show.legend=FALSE) +
  scale_y_log10(labels = scales::comma) +
  scale_x_log10(labels = scales::comma) +
  xlab("total tests (log scale)") + ylab("total cases (log scale)") +
  ggtitle(title) +
  theme_minimal() + theme(legend.position = "right")
```








